tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2024-03-13 paper.tex
Minor wording tweaks

% \journaltitle{Bioinformatics}
\journaltitle{Preprint}
% \DOI{DOI HERE}
\copyrightyear{2023}
% \pubyear{2019}
% \access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

@@

An important property of these population genetic simulation methods is that
they output ancestral recombination graphs (ARGs)
rather than sample genotypes.
ARGs encode the interwoven paths
of genetic inheritance arising from
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and contain rich detail about ancestral processes.
Recent breakthroughs in inferrence methods have made it possible
to estimate ARGs at biobank scale~\citep{kelleher2019,zhang2023},
and there is now intense interest in their practical
application~\citep{lewanski2024era,brandt2024promise}.
Statistical genetics has been a particular focus,
and ARG-based methods have been shown
to detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023}; to
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and to provide a sparse and efficient model of linkage disequilibrium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

@@

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using familiar Python data science tools.

>>>>

% \journaltitle{Bioinformatics}
\journaltitle{Preprint}
% \DOI{DOI HERE}
\copyrightyear{2024}
% \pubyear{2019}
% \access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

@@

An important property of these population genetic simulation methods is that
they output ancestral recombination graphs (ARGs)
rather than sample genotypes.
ARGs encode the interwoven paths
of genetic inheritance caused by
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination \citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and contain rich detail about ancestral processes.
Recent breakthroughs in inferrence methods have made it possible
to estimate ARGs at biobank scale~\citep{kelleher2019,zhang2023},
and there is now intense interest in their practical
application~\citep{lewanski2024era,brandt2024promise}.
Statistical genetics has been a particular focus,
and ARG-based methods have been shown
to detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023}; to
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and to provide a sparse and efficient model of linkage disequilibrium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

@@

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, representing the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using standard Python data science tools.
